---
title: 'NYC Real Estate Transactions'
author: "Quentin McTeer"
date: "5/02/2021"
output:
  html_document:
    toc: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
  '': default
---

# Preprocessing

```{r Preprocessing, message=FALSE, warning=FALSE}
library(tidyverse) 
library(broom)
library(modelsummary)
library(kableExtra)
library(cowplot)
library(qwraps2)
library(gtsummary)
library(skimr)
library(readxl)
library(fixest)
library(tidymodels)
library(pROC)
library(corrplot)
library(Hmisc)
library(huxtable)
options(qwraps2_markup = "markdown") 
options(scipen = 999) # load packages and set options 

nyc <- read_csv("nyc-rolling-sales.csv")
inc <- read_xlsx("inc.xlsx")  # read in data 
```

```{r skim, message=FALSE, warning=FALSE}
 # skim(nyc) # evaluate data structure and missing values
```

```{r rename , message=FALSE, warning=FALSE}
nyc <- nyc %>% 
  rename("id" = X1,
         "borough" = BOROUGH,
         "neighborhood" = NEIGHBORHOOD,
         "bld_clss_cat" = `BUILDING CLASS CATEGORY`,
         "tax_clss_pres" = `TAX CLASS AT PRESENT`,
         "block" = BLOCK,
         "lot" = LOT, 
         "easement" = `EASE-MENT`,
         "bld_clss_pres" = `BUILDING CLASS AT PRESENT`,
         "address" = ADDRESS,
         "apt_no" = `APARTMENT NUMBER`,
         "zip" = `ZIP CODE`,
         "resid_units" = `RESIDENTIAL UNITS`,
         "commer_units" = `COMMERCIAL UNITS`,
         "ttl_units" = `TOTAL UNITS`,
         "lnd_sqft" = `LAND SQUARE FEET`,
         "gross_sqft" = `GROSS SQUARE FEET`,
         "year_built" = `YEAR BUILT`,
         "tax_clss_sale" = `TAX CLASS AT TIME OF SALE`,
         "bld_clss_sale" = `BUILDING CLASS AT TIME OF SALE`,
         "price" = `SALE PRICE`,
         "date" = `SALE DATE`) # rename columns 
```

# Create Analytic Dataset 

## Merge Datasets

```{r merge , message=FALSE, warning=FALSE}
nyc <- left_join(x = nyc, y = inc, by = "borough") # merge in median income, crime rates, and percent white by borough data 
```

## Deal With Missing Values 
```{r missing values, message=FALSE, warning=FALSE}
nyc <- nyc %>% 
  mutate(lnd_sqft = ifelse(lnd_sqft == "-", NA, lnd_sqft),
         gross_sqft = ifelse(gross_sqft == "-", NA, gross_sqft),
         price = ifelse(price == "-", NA, price),
         year_built = ifelse(year_built == 0, NA, year_built)) # create NA values for missing data 

nyc <- nyc %>% 
  select(-easement, -apt_no, -lot, -bld_clss_pres, -tax_clss_pres, -block, -zip) # eliminate easement and apartment number columns, they will not help us 

nyc <- na.omit(nyc) # omit missing values, imputation won't make sense for many missing values in the dataset

```

## Change data structure
```{r structure , message=FALSE, warning=FALSE}
nyc <- nyc %>% 
  mutate(bld_clss_cat = as.factor(bld_clss_cat),
         lnd_sqft = as.numeric(lnd_sqft),
         gross_sqft = as.numeric(gross_sqft),
         bld_clss_sale = as.factor(bld_clss_sale),
         price = as.numeric(price),
         borough = as.factor(borough),
         tax_clss_sale = as.factor(tax_clss_sale),
         white = as.numeric(white)) # coerce columns to fit more accurate data structure 
```

## Feature Engineering
```{r deed transfer, message=FALSE, warning=FALSE}
nyc %>% 
  filter(price < 1500) %>% 
  count() # check number of deed transfers 

nyc <- nyc %>% 
  mutate(deed_trans = ifelse(price < 1500, "TRUE", "FALSE" )) # generate deed transfer indicator variable 
```

```{r build class cat, message=FALSE, warning=FALSE}
nyc %>% 
  group_by(bld_clss_cat) %>% 
  count() %>% 
  arrange(desc(n)) %>%
  kable()

nyc <- nyc %>% 
  mutate(bld_clss_cat = fct_lump(bld_clss_cat, prop = 0.045)) # shrink building class category to seven factors
```

```{r residential only, message=FALSE, warning=FALSE}
nyc <- nyc %>% 
  filter(bld_clss_cat != "Other") # include only residential property transactions 
```

```{r build class sale, message=FALSE, warning=FALSE}
nyc %>% 
  group_by(bld_clss_sale) %>% 
  count() %>% 
  arrange(desc(n)) %>%
  kable() # check number of categories and frequency of each category

nyc <- nyc %>% 
  mutate(bld_clss_sale = fct_lump(bld_clss_sale, prop = 0.04))  # shrink building class at time of sale to ten factors
```

```{r mean gross sqft, message=FALSE, warning=FALSE}
gen <- nyc %>% 
  group_by(borough) %>%
  summarise(mean_sqft = mean(gross_sqft))# create mean square feet by borough variable 

nyc <- left_join(x = nyc, y = gen, by = "borough") # merge the variable into the original data set
```

```{r borough, message=FALSE, warning=FALSE}
nyc <- nyc %>% 
  mutate(borough = ifelse(borough == 1, "Manhattan", borough),
         borough = ifelse(borough == 2, "Bronx", borough),
         borough = ifelse(borough == 3, "Brooklyn", borough),
         borough = ifelse(borough == 4, "Queens", borough),
         borough = ifelse(borough == 5, "Staten Island", borough)) # edit borough variable to display borough name instead of 1-5 value 
```

```{r date variables, message=FALSE, warning=FALSE}
nyc <- separate(nyc, date, c("year", "month", "day"), sep = "-") # create separate year, month and day variables from data column 

nyc$date <- paste(nyc$year, nyc$month, nyc$day, sep="/") # Recreate date variable 

nyc$date <- as.Date(nyc$date) # Read date variable as date 
```

# Exploratory Data Analysis 

## Visualizations 
```{r Building Sale Price Distribution, message=FALSE, warning=FALSE}
nyc %>% 
  filter(price < 5000000 & deed_trans == FALSE) %>%
  ggplot(nyc, mapping = aes(x = price)) +
  geom_histogram(bins = 12, fill = "#90BDDD", color = "white") +
  ylab("Number of Transactions") + 
  xlab("Sale Price (in 000's)") + 
  theme_cowplot(12) # plot sale price distribution 
```

```{r Real Estate Sales by Borough: Sep 2016-Sep 2017, message=FALSE, warning=FALSE}
nyc %>%
  group_by(borough) %>%
  summarise(n = n()) %>%
  ggplot() + 
  geom_bar(aes(x = borough, y = n), stat = "identity", fill = "#90BDDD") +
  ylab("Number of Transactions") + 
  xlab("Borough") + 
  theme_cowplot(12) # plot real estate sales by borough 
```

```{r Average Sale Price By Building Type, message=FALSE, warning=FALSE}
nyc %>%
  filter(deed_trans == FALSE) %>%
  group_by(bld_clss_cat) %>%
  summarise(avg.price = mean(price)) %>%
  ggplot() + 
  geom_bar(aes(x = avg.price/1000, y = bld_clss_cat), stat = "identity", fill = "#90BDDD") +
  ylab("Building Class Category") + 
  xlab("Price (000's of $)") + 
  theme_cowplot(12) # plot average sale price price by building type 


```

```{r Average Building Sale Price By Building Class at Sale, message=FALSE, warning=FALSE}
nyc %>%
  filter(deed_trans == FALSE) %>%
  group_by(bld_clss_sale) %>%
  summarise(avg.price = mean(price)/1000) %>%
  ggplot() + 
  geom_bar(aes(x = avg.price, y = bld_clss_sale), stat = "identity", fill = "#90BDDD") +
  ylab("Building Class at Sale") + 
  xlab("Price in 000's in $") + 
  theme_cowplot(12) # plot average sale price by building class at sale 
```

```{r Seasonal Trends in New York City Real Estate Sales, message=FALSE, warning=FALSE}
nyc %>%
  filter(deed_trans == FALSE) %>%
  group_by(month) %>%
  count() %>%
  ggplot() + 
  geom_point(aes(x = month, y = n)) + 
  geom_line(aes(x = month, y = n, group = 1)) +
  geom_vline(aes(xintercept = "09"), color = "#90BDDD" ) +
  ylab("Number of Transactions") + 
  xlab("Month") + 
  theme_cowplot(12) # plot monthly trends in NYC real estate transactions 
```

```{r Impact of Building Age on Sales Price, message=FALSE, warning=FALSE}
nyc %>%
  filter(deed_trans == FALSE & price < 5000000) %>%
  group_by(year_built) %>%
  summarise(avg.price = mean(price)) %>%
  ggplot() + 
  geom_point(aes(x = year_built, y = avg.price)) +
  ylab("Average Sale Price") + 
  xlab("Year Built") + 
  geom_smooth(aes(x = year_built, y = avg.price), color = "#90BDDD", method = "lm") + 
  theme_cowplot(12) # plot year built against sale price 

```

```{r Relationship Between Building Tax Class and Sale Price, message=FALSE, warning=FALSE}
nyc %>%
  group_by(tax_clss_sale) %>%
  summarise(avg.price = mean(price)) %>%
  ggplot() + 
  geom_bar(aes(x = avg.price/1000, y = tax_clss_sale), stat = "identity", fill = "#90BDDD") +
  xlab("Average Sale Price (000's of $'s) ") + 
  ylab("Building & Building Units: Tax Class") + 
  theme_cowplot(12) # plot bar chart of tax class on sale price 
```

```{r Impact of Square Footage on Sale Price, message=FALSE, warning=FALSE}
nyc %>%
  filter(deed_trans == FALSE) %>%
  filter(gross_sqft > 0 & gross_sqft < 10000 & price < 10000000) %>%
  group_by(gross_sqft) %>% 
  summarise(avg.price = mean(price)) %>%
  ggplot() + 
  geom_point(aes(x = gross_sqft, y = avg.price/1000)) +
  theme_cowplot(12) +
  ylab("Average Sale Price (000's of $)") + 
  xlab("Gross Square Feet") + 
  geom_smooth(aes(x = gross_sqft, y = avg.price/1000), color = "#90BDDD", method = "lm") # plot gross square feet against sale price 
```

## Summary Statistics 
```{r Summary Stats Deed Trans, message=FALSE, warning=FALSE}
nyc %>%
  select(deed_trans, resid_units, commer_units, ttl_units, lnd_sqft, gross_sqft, year_built, price, borough, tax_clss_sale) %>%
  tbl_summary(
    by = deed_trans,
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    digits = all_continuous() ~ 2,
  ) %>%
  as_hux_table() %>%
  theme_plain # create summary statistics table segmented by deed transfer 

```


```{r summary stats borough, message=FALSE, warning=FALSE}
tb_hux <- nyc %>%
  select(income, crime, white, resid_units, commer_units, ttl_units, lnd_sqft, gross_sqft, year_built, price, borough) %>%
  tbl_summary(
    by = borough,
    statistic = list(all_continuous() ~ "{mean}"),
    digits = all_continuous() ~ 2,
    type = list(income ~ 'continuous',
                crime ~ 'continuous',
                white ~ 'continuous')) %>%
  as_hux_table # create summary statistics table 
  
tb_hux %>%
  set_width(0.9) %>%
set_font_size(10) %>%
  theme_plain() %>%
set_position("center") # report summary statistics table
  
```

```{r correlation matrix, message=FALSE, warning=FALSE}
res2 <- cor((nyc[,c("income", "crime", "white", "resid_units", "commer_units", "ttl_units", "lnd_sqft", "gross_sqft", "price")])) # create correlation matrix 

res2 %>% 
  kbl(digits = 3, booktabs = T) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = c("striped", "hold_position")) # report correlation matrix 
```

# Regression and Machine Learning Models

## Causal Inference Model
```{r linear regression model, message=FALSE, warning=FALSE}
nyc_pred <- nyc %>% 
  filter(deed_trans == FALSE) %>%
  select(-neighborhood, -id, -address, -day, -date, -mean_sqft, -year,
         -deed_trans, -white, -ttl_units, -crime) %>% 
  mutate(month = as.factor(month),
         borough = as.factor(borough)) # create prediction data set from analytic data set for linear regression and 

borough_model <- feols(price ~ borough + resid_units + commer_units + lnd_sqft + tax_clss_sale                                     + year_built + year_built^2 | month + bld_clss_cat + bld_clss_sale, 
                               data = nyc_pred) # build causal inference model 

tidy(borough_model, se = "white") %>%
  kbl(digits = 3, booktabs = T) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = c("striped", "hold_position")) # create table 

glance(borough_model)

RSS <- c(crossprod(borough_model$residuals))
MSE <- RSS / length(borough_model$residuals)
RMSE <- sqrt(MSE)
RMSE # pull out metrics 
```

## Random Forest Decision Tree Ensemble Prediction Model
```{r random forest regression, message=FALSE, warning=FALSE}
set.seed(555)

nyc_split <- initial_split(nyc_pred)

nyc_train <- training(nyc_split)
nyc_test <- testing(nyc_split) # create test and training dataset 

nyc_wf <- workflow() %>%
  add_formula(price ~ .)

rf_spec <- rand_forest(
  mtry = 6,
  trees = 1000,
 min_n = 5) %>%
 set_mode("regression") %>% 
  set_engine(engine = "ranger") 

  rf_final <- nyc_wf %>% 
 add_model(rf_spec) %>% 
 last_fit(nyc_split) # train model and gather test results

collect_metrics(rf_final) # report metrics 

```


## Logistic Regression Prediction Model
```{r logistic regression classification, message=FALSE, warning=FALSE}

set.seed(1234)
nyc_bin <- nyc %>% 
 select(-neighborhood, -id, -address, -day, -date, -mean_sqft, -year, -crime, -price, -income, -white) %>% 
 mutate(month = as.factor(month),
        deed_trans = as.factor(deed_trans),
        borough = as.factor(borough)) # create dataset classification models 

nyc_split <- initial_split(nyc_bin)

nyc_train1 <- training(nyc_split)
nyc_test1 <- testing(nyc_split) # create training/test data

 nyc_wf1 <- workflow() %>%
 add_formula(deed_trans ~ .)

 log_spec <- logistic_reg() %>% 
  set_engine(engine = "glm") # build model 

 log_final <- nyc_wf1 %>% 
 add_model(log_spec) %>% 
 last_fit(nyc_split) # run model on train/test data 

collect_metrics(log_final) # report metrics

```

## Random Forest Classification Decision Tree Ensemble Model
```{r random forest classification, message=FALSE, warning=FALSE}
 set.seed(542)
rf_spec1 <- rand_forest(
 mtry = 8,
 trees = 1000,
 min_n = 8) %>%
 set_mode("classification") %>% 
 set_engine(engine = "ranger") # create model 

 rf_final1 <- nyc_wf1 %>% 
 add_model(rf_spec1) %>% 
 last_fit(nyc_split) # run model on training and test data

 collect_metrics(rf_final1) # report metrics 

```


